---
title: "Hybrid imputation with RF + LCMD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hybrid imputation with RF + LCMD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Step 4: Hybrid Imputation (Random Forest + LCMD)

```{r eval=FALSE}
imputed_results <- OmicsProcessing::hybrid_imputation(
  log_transformed_df,
  target_cols = "@",
  method = c("RF-LCMD"),
  oobe_threshold = 0.1
)
imputed_df <- imputed_results$hybrid_rf_lcmd
```

`hybrid_imputation()` combines two complementary strategies:

* **Random Forest (RF)** via `missForest::missForest()` for features that
  are well-predicted by other variables (low OOBE).
* **Left-Censored Missing Data (LCMD)** via `imputeLCMD::impute.MAR.MNAR()`
  for features with higher OOBE.

The function fits RF per feature, uses the out-of-bag error (OOBE) to
decide whether to keep the RF estimate or switch that feature to LCMD,
and returns:

* `hybrid_rf_lcmd`: the combined result.
* `rf` / `lcmd`: per-method outputs.
* `oob`: OOBE values (helpful for diagnostics).

See the full reference: [`hybrid_imputation()`](../reference/hybrid_imputation.html).

### Handling QC samples with `is_qc`

Rows flagged as `TRUE` by `is_qc` are excluded from the hybrid imputation
calculations. The rationale is that QC samples are repeatedly measured
throughout the laboratory workflow and are intended to capture technical
variability in the measurement process. Excluding them prevents these rows
from influencing the RF and LCMD imputation models fitted to the biological
samples.

If QC samples contain missing values that you also want to impute, there
are two main options.

#### A. Impute QC samples together with the rest of the data

To include all rows in the hybrid imputation process, leave `is_qc` as
`NULL`, or provide a logical vector of the same length as the number of
rows in the data, with all values set to `FALSE`.

```{r eval=FALSE}
imputed_results <- OmicsProcessing::hybrid_imputation(
  log_transformed_df,
  target_cols = "@",
  method = "RF-LCMD",
  oobe_threshold = 0.1,
  is_qc = NULL
)

is_qc_all_false <- rep(FALSE, nrow(log_transformed_df))

imputed_results <- OmicsProcessing::hybrid_imputation(
  log_transformed_df,
  target_cols = "@",
  method = "RF-LCMD",
  oobe_threshold = 0.1,
  is_qc = is_qc_all_false
)
```

#### B. Impute QC samples using only QC sample values

If you want to impute QC samples separately, using only information from
other QC samples, you can invert the `is_qc` vector. This excludes all
non-QC rows from the imputation calculations and applies the pipeline only
to the QC samples.

```{r eval=FALSE}
is_qc <- sample_metadata$is_qc

imputed_qc_results <- OmicsProcessing::hybrid_imputation(
  log_transformed_df,
  target_cols = "@",
  method = "RF-LCMD",
  oobe_threshold = 0.1,
  is_qc = !is_qc
)

imputed_qc_df <- imputed_qc_results$hybrid_rf_lcmd
```


### Customising the RF and LCMD controls

You can tweak both engines via control lists:

```{r eval=FALSE}
my_control_RF <- list(
  parallelize = "no",
  mtry = floor(sqrt(length(target_cols))),
  ntree = 100,
  maxiter = 10,
  variablewise = TRUE,
  verbose = TRUE,
  n_cores = parallel::detectCores()
)

my_control_LCMD <- list(
  method.MAR = "KNN",
  method.MNAR = "QRILC"
)

df_rf_lcmd_hybrid <- OmicsProcessing::hybrid_imputation(
  log_transformed_df,
  target_cols = "@",
  method = c("RF-LCMD"),
  oobe_threshold = 0.1,
  control_LCMD = my_control_LCMD,
  control_RF = my_control_RF
)
```

### Parallelising the RF step (`missForest`)

`missForest` can run in parallel when you register a **foreach**
backend and set `parallelize`:

```{r eval=FALSE}
library(doParallel)

n_cores <- parallel::detectCores(logical = FALSE)
cl <- parallel::makeCluster(n_cores)
doParallel::registerDoParallel(cl)

ctrl_parallel_RF <- list(
  parallelize = "variables", # or "forests"
  mtry = floor(sqrt(length(target_cols))),
  ntree = 200,
  maxiter = 10,
  variablewise = TRUE,
  verbose = TRUE
)

imputed_parallel <- OmicsProcessing::hybrid_imputation(
  log_transformed_df,
  target_cols = "@",
  method = "RF-LCMD",
  oobe_threshold = 0.1,
  control_RF = ctrl_parallel_RF
)

parallel::stopCluster(cl)
doParallel::registerDoSEQ()
```

Guidance:

* Use `parallelize = "variables"` for many features; `"forests"` spreads
  trees instead.
* Keep `ntree` reasonable when parallelising to avoid memory pressure.


Tips:

* Keep `target_cols` explicit when possible for clarity; `"@"` will use
  all feature columns resolved via `resolve_target_cols()`.
* Inspect `imputed_results$oob` to confirm the RF ↔ LCMD split aligns
  with your expectations.
* For very wide matrices, tune `ntree`, `mtry`, or the number of worker
  cores to balance runtime and stability.
