---
title: "Hybrid imputation with RF + LCMD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hybrid imputation with RF + LCMD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Step 4: Hybrid Imputation (Random Forest + LCMD)

```{r eval=FALSE}
imputed_results <- OmicsProcessing::hybrid_imputation(
  log_transformed_df,
  target_cols = "@",
  method = c("RF-LCMD"),
  oobe_threshold = 0.1
)
imputed_df <- imputed_results$hybrid_rf_lcmd
```

`hybrid_imputation()` combines two complementary strategies:

* **Random Forest (RF)** via `missForest::missForest()` for features that
  are well-predicted by other variables (low OOBE).
* **Left-Censored Missing Data (LCMD)** via `imputeLCMD::impute.MAR.MNAR()`
  for features with higher OOBE.

The function fits RF per feature, uses the out-of-bag error (OOBE) to
decide whether to keep the RF estimate or switch that feature to LCMD,
and returns:

* `hybrid_rf_lcmd`: the combined result.
* `rf` / `lcmd`: per-method outputs.
* `oob`: OOBE values (helpful for diagnostics).

See the full reference: [`hybrid_imputation()`](../reference/hybrid_imputation.html).

### Customising the RF and LCMD controls

You can tweak both engines via control lists:

```{r eval=FALSE}
my_control_RF <- list(
  parallelize = "no",
  mtry = floor(sqrt(length(target_cols))),
  ntree = 100,
  maxiter = 10,
  variablewise = TRUE,
  verbose = TRUE,
  n_cores = parallel::detectCores()
)

my_control_LCMD <- list(
  method.MAR = "KNN",
  method.MNAR = "QRILC"
)

df_rf_lcmd_hybrid <- OmicsProcessing::hybrid_imputation(
  log_transformed_df,
  target_cols = "@",
  method = c("RF-LCMD"),
  oobe_threshold = 0.1,
  control_LCMD = my_control_LCMD,
  control_RF = my_control_RF
)
```

### Parallelising the RF step (`missForest`)

`missForest` can run in parallel when you register a **foreach**
backend and set `parallelize`:

```{r eval=FALSE}
library(doParallel)

n_cores <- parallel::detectCores(logical = FALSE)
cl <- parallel::makeCluster(n_cores)
doParallel::registerDoParallel(cl)

ctrl_parallel_RF <- list(
  parallelize = "variables", # or "forests"
  mtry = floor(sqrt(length(target_cols))),
  ntree = 200,
  maxiter = 10,
  variablewise = TRUE,
  verbose = TRUE
)

imputed_parallel <- OmicsProcessing::hybrid_imputation(
  log_transformed_df,
  target_cols = "@",
  method = "RF-LCMD",
  oobe_threshold = 0.1,
  control_RF = ctrl_parallel_RF
)

parallel::stopCluster(cl)
doParallel::registerDoSEQ()
```

Guidance:

* Use `parallelize = "variables"` for many features; `"forests"` spreads
  trees instead.
* Keep `ntree` reasonable when parallelising to avoid memory pressure.


Tips:

* Keep `target_cols` explicit when possible for clarity; `"@"` will use
  all feature columns resolved via `resolve_target_cols()`.
* Inspect `imputed_results$oob` to confirm the RF ↔ LCMD split aligns
  with your expectations.
* For very wide matrices, tune `ntree`, `mtry`, or the number of worker
  cores to balance runtime and stability.
