---
title: "Normalisation plot comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Normalisation plot comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Compare run-order plots before and after normalisation

Use `plot_omics_distributions()` to visualise feature intensities
across run order before and after normalisation. The plot is useful for
checking whether QC drift and plate or batch effects are reduced after
processing.

```{r eval=FALSE}
target_features <- c("feature_1", "feature_2", "feature_3")

p <- OmicsProcessing:::plot_omics_distributions(
  df = df_pre_normalised,
  df_comp = df_normalised,
  target_cols = target_features,
  run_order = "run_order",
  is_qc = "is_qc",
  batch = "batch",
  plate = "plate",
  title_ref = "Before normalisation",
  title_comp = "After normalisation",
  point_size = 1.2
)

p
```

The reference data frame (`df`) and comparison data frame (`df_comp`)
should contain the same feature columns and run-order variable. When
`df_comp` is supplied, the function returns side-by-side plots with a
shared legend so you can compare the effect of normalisation directly.

### Minimal input requirements

- `target_cols`: character vector of feature columns to facet.
- `run_order`: column containing injection or acquisition order.
- `is_qc`: optional logical QC indicator column; if omitted, all samples
  are treated as non-QC.
- `batch`: optional batch column used to draw vertical batch boundary
  lines.
- `plate`: optional plate column used to colour points.

### Typical workflow

This plot is most useful after you have:

1. filtered features and samples,
2. imputed missing values if needed, and
3. created a normalised data frame, for example with
   `normalise_SERRF()`.

```{r eval=FALSE}
df_normalised <- OmicsProcessing::normalise_SERRF(
  df_imputed,
  target_cols = target_features,
  is_qc = df_imputed$is_qc,
  strata_col = "batch"
)

OmicsProcessing:::plot_omics_distributions(
  df = df_imputed,
  df_comp = df_normalised,
  target_cols = target_features,
  run_order = "run_order",
  is_qc = "is_qc",
  batch = "batch",
  plate = "plate"
)
```

### Interpretation tips

- QC samples should show smoother behaviour across run order after
  normalisation if technical drift has been reduced.
- Plate colouring can help reveal residual plate-specific structure.
- Batch boundary lines are useful for checking whether step changes
  remain after correction.

If you are using SERRF for correction, see the related vignette:
[Batch correction using SERRF](../articles/serrf-normalisation.html).
